Estimating species interactions in large, abiotically structured communities with Markov networks and stochastic approximation

### 2015-08-20 stochastic/stochastic-recovery.Rmd
chapter 3 stuff

```{r}
set.seed(1)
n_spp = 250
n_loc = 2500
n_env = 3
n_gibbs = 1000

true_beta_vec = rnorm(
  choose(n_spp, 2), 
  sd = abs(rnorm(choose(n_spp, 2))) / 5
)

@@

true_intercepts = rnorm(n_spp)

true_env = matrix(rnorm(n_loc * n_env, 0), nrow = n_loc)

true_alpha_env = matrix(rnorm(n_spp * n_env, 0, 0.5), nrow = n_env)

true_alpha = true_env %*% true_alpha_env
```

@@

alpha_env = delta_alpha_env = matrix(0, nrow = n_env, ncol = n_spp)
alpha_species = delta_alpha_species = qlogis(colMeans(y))
beta = delta_beta = matrix(0, nrow = n_spp, ncol = n_spp)
alpha = matrix(0, nrow = n_spp, ncol = n_spp) # no delta alpha to initialize
                                              # b/c alpha not optimized directly
```

```{r}
initial_learning_rate = 0.1
momentum = 0.9

pb <- progress_bar$new(
  format = "  Fitting [:bar] :percent eta: :eta",
  total = n_gibbs,
  clear = FALSE
)

for(i in 1:n_gibbs){
  pb$tick()

@@

  # Update learning rate
  learning_rate = initial_learning_rate * 1000 / (999 + i)

@@

  # Calculate the gradient with respect to alpha and beta
  delta_stats = y_stats - y_sim_stats
  beta_grad = (delta_stats - 10 * beta) / 10 / n_loc # Subtract beta as weight decay

  alpha_species_grad = diag(beta_grad) / n_loc
  diag(beta_grad) = 0
  alpha_env_grad = (y_env_stats - y_sim_env_stats - alpha_env)  / n_loc

  # Calculate parameter updates
  delta_beta = beta_grad * learning_rate + momentum * delta_beta
  delta_alpha_species = alpha_species_grad * learning_rate # no momentum
  delta_alpha_env = alpha_env_grad * learning_rate + momentum  * delta_alpha_env

@@

  if(i %% 10 == 0){
    plot(beta[upper.tri(beta)], true_beta[upper.tri(beta)],
         pch = 16, cex = .5, col = "#00000050",
         xlab = "Estimated interaction strength",
         ylab = "\"True\" interaction strength",
         main = paste(
           "R^2 =", 
           signif(cor(beta[upper.tri(beta)], true_beta[upper.tri(beta)])^2, 2)
          )
    )
    abline(v = 0, h = 0)
  }
}
```

```{r}
pdf("poster-beta.pdf", width = 4, height = 5)
plot(beta[upper.tri(beta)], true_beta[upper.tri(beta)],
     pch = 16, cex = .4, col = "#00000020",
     xlab = "Estimated interaction strength",
     ylab = "\"True\" interaction strength",
     main = paste0("Recoverd pairwise interactions", "\nN = ", choose(n_spp, 2)),
     las = 1,
     cex.axis = 0.75
    )
text(
  -0.7, 
  1,
  paste(
    "R^2 =", 
    signif(cor(beta[upper.tri(beta)], true_beta[upper.tri(beta)])^2, 2)
  )
)

dev.off()
```

```{r}
pdf("poster-alpha.pdf", width = 4, height = 5)
plot(alpha_env, true_alpha_env,
     pch = 16, cex = .4, col = "#00000090",
     xlab = "Estimated environmental coefficient",
     ylab = "\"True\" environmental coefficient",
     main = paste0("Recoverd environmental coefficients", "\nN = ", length(alpha_env)),
     las = 1,
     cex.axis = 0.75
)
text(
  -1, 
  1,
  paste(
    "R^2 =", 
    signif(cor(c(alpha_env), c(true_alpha_env))^2, 2)
  )
)
dev.off()

@@

>>>>

```{r}
set.seed(1)
n_spp = 250
n_loc = 2500
n_env = 5
n_gibbs = 5000

type_frequencies = rmultinom(1, size = choose(n_spp, 2), prob = c(.2, .5, .3))

true_beta_vec = sample(
  c(
    rnorm(type_frequencies[[1]], 0, 2), 
    rnorm(type_frequencies[[2]], 0, .2), 
    rnorm(type_frequencies[[3]], -.5, .5)
  )
)

plot(density(true_beta_vec))

@@

true_env = matrix(rnorm(n_loc * n_env, 0, 2.5), nrow = n_loc)

true_alpha_env = matrix(rnorm(n_spp * n_env, 0, 1), nrow = n_env)

true_alpha_species = rnorm(n_spp, 2)

true_alpha = true_env %*% true_alpha_env %plus% true_alpha_species
```

@@

alpha_env = delta_alpha_env = matrix(0, nrow = n_env, ncol = n_spp)
alpha_species = qlogis((colSums(y) + 1) / (nrow(y) + 2))
delta_alpha_species = rep(0, n_spp)
beta = delta_beta = matrix(0, nrow = n_spp, ncol = n_spp)
alpha = matrix(0, nrow = n_spp, ncol = n_spp) # no delta alpha to initialize
                                              # b/c alpha not optimized directly
```

```{r}
# Are there lurking factors of two because I use the whole symmetric matrix?
beta_prior = rosalia::make_logistic_prior(scale = 0.5)$log_grad
alpha_env_prior = rosalia::make_logistic_prior(scale = 2)$log_grad 
alpha_species_prior = rosalia::make_logistic_prior(scale = 2)$log_grad
```

```{r}
initial_learning_rate = 1
maxit = 50000
start_time = as.integer(Sys.time())

pb <- progress_bar$new(
  format = "  Fitting [:bar] :percent eta: :eta",
  total = maxit,
  clear = FALSE
)

r2s = numeric(maxit)
times = integer(maxit)

for(i in 1:maxit){
  pb$tick()

@@

  # Update learning rate and momentum
  learning_rate = initial_learning_rate * 1000 / (998 + 1 + i)
  momentum = .9 * (1 - 1/(.1 * i + 2))

@@

  # Calculate the gradient with respect to alpha and beta
  stats_difference = y_stats - y_sim_stats
  beta_grad = (stats_difference + beta_prior(beta)) / n_loc

  alpha_species_grad = (diag(stats_difference) + alpha_species_prior(alpha_species)) / n_loc
  diag(beta_grad) = 0
  y_env_difference = y_env_stats - y_sim_env_stats
  alpha_env_grad = (y_env_difference + alpha_env_prior(alpha_env))  / n_loc

  # Calculate parameter updates
  delta_beta = beta_grad * learning_rate / 10 + 
    momentum * delta_beta
  delta_alpha_species = alpha_species_grad * learning_rate + 
    momentum  * delta_alpha_species
  delta_alpha_env = alpha_env_grad * learning_rate + 
    momentum  * delta_alpha_env

@@

  r2s[i] = cor(true_beta[upper.tri(true_beta)], beta[upper.tri(beta)])^2
  times[i] = as.integer(Sys.time()) - start_time
}
```

```{r}
library(cowplot)
library(ggplot2)
out = plot_grid(
  ggplot(NULL, aes(x = beta[upper.tri(beta)], y = true_beta[upper.tri(beta)])) + 
    stat_binhex() + 
    xlab("Estimated coefficient value") + 
    ylab("\"True\" coefficient value") + 
    stat_hline(yintercept = 0, size = 1/8) + 
    stat_vline(xintercept = 0, size = 1/8) + 
    scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "log10") +
    stat_abline(intercept = 0, slope = 1, size = 1/2) + 
    coord_equal(),  
  ggplot(NULL, aes(x = c(alpha_env), y = c(true_alpha_env))) + 
    stat_binhex() + 
    xlab("Estimated coefficient value") + 
    ylab("\"True\" coefficient value") + 
    stat_hline(yintercept = 0, size = 1/8) + 
    stat_vline(xintercept = 0, size = 1/8) + 
    scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "log10") +
    stat_abline(intercept = 0, slope = 1, size = 1/2) + 
    coord_equal(),  
  nrow = 1,
  labels = c("A. Pairwise biotic coefficients (n = 31125)", "B. Abiotic coefficients (n = 1250)")
)
save_plot("stochastic/estimates.pdf", out, base_aspect_ratio = 2, base_height = 6)

@@

## Small landscape

```{r}
small_n_spp = 20
small_n_loc = 500
small_true_beta = true_beta[1:small_n_spp, 1:small_n_spp]
small_true_alpha = true_alpha_species - 2
small_y = matrix(0.5, nrow = small_n_loc, ncol = small_n_spp)

for(i in 1:n_gibbs){
  for(j in sample.int(small_n_spp)){
    small_y[,j] = rbern(logistic(small_true_alpha[j] + small_y %*% small_true_beta[ , j]))
  }
}

```{r}
exact_time = system.time({
  exact = rosalia(small_y, prior = make_logistic_prior(scale = 1), maxit = 500)
})

mle_beta = exact$beta
```

```{r}
maxit_small = 50000
start_time_small = as.numeric(Sys.time())

beta_prior_small = rosalia::make_logistic_prior(scale = 1)$log_grad
alpha_species_prior_small = rosalia::make_logistic_prior(scale = 1)$log_grad

alpha_small = qlogis((colSums(small_y) + 1) / (nrow(small_y) + 2))
delta_alpha_small = rep(0, small_n_spp)
beta_small = delta_beta_small = matrix(0, nrow = small_n_spp, ncol = small_n_spp)

y_sim_small = matrix(0.5, nrow = small_n_loc, ncol = small_n_spp)

y_stats = crossprod(small_y)

mses = numeric(maxit_small)
small_times = numeric(maxit_small)

for(i in 1:maxit_small){
  ##############################
  # Gibbs sampling for predicted species composition
  ##############################

  # Sample entries in y_sim from their conditional distribution (Gibbs sampling)
  for(j in sample.int(small_n_spp)){
    y_sim_small[,j] = rbern(logistic(alpha_small[j] + y_sim_small %*% beta_small[ , j]))
  }

  # Update learning rate and momentum
  learning_rate = initial_learning_rate * 1000 / (998 + 1 + i)
  momentum = .9 * (1 - 1/(.1 * i + 2))

  # Calculate sufficient statistics
  y_sim_stats = crossprod(y_sim_small)

  # Calculate the gradient with respect to alpha and beta
  stats_difference = y_stats - y_sim_stats
  beta_grad = (stats_difference + beta_prior_small(beta_small)) / small_n_loc

  alpha_species_grad = (diag(stats_difference) +
                          alpha_species_prior_small(alpha_small)) / small_n_loc
  diag(beta_grad) = 0

  # Calculate parameter updates
  delta_beta = beta_grad * learning_rate + 
    momentum * delta_beta_small
  delta_alpha_species = alpha_species_grad * learning_rate + 
    momentum  * delta_alpha_small

  beta_small = beta_small + delta_beta
  alpha_small = alpha_small + delta_alpha_species

  mses[i] = mean((beta_small[upper.tri(beta_small)] - exact$beta[upper.tri(exact$beta)])^2)
  small_times[i] = as.numeric(Sys.time()) - start_time_small
}
```

## Plotting results

```{r}
pdf("stochastic/convergence.pdf", width = 10, height = 7)
par(mfrow = c(1, 2), las = 1)

plot(small_times, mses, type = "l", ylab = "mean square deviation from MLE\n(log scale)", xlab = "time (seconds)", main = "A. Convergence to known MLE\n(small network)", yaxs = "i", bty = "l", log = "y", xaxs = "i", lwd = 2, col = "blueviolet", xlim = range(c(-.1, small_times)))

plot(c(0, times / 60 / 60), c(0, r2s), type = "l", xlab = "time (hours)", ylab = "R^2",ylim = c(0, max(r2s) * 1.04), main = "B. Apparent convergence to unknown MLE\n(large network)", yaxs = "i", xaxs = "i", lwd = 2, xlim = range(c(-.05, times / 60 / 60)), bty = "l", col = "blueviolet")
abline(h = max(r2s), col = "#00000080", lty = 2, lwd = 2)
dev.off()
```
